---
title: "A review of data visualizations"
author: "Philippe Fullsack and Albert Einstain"
date: "2025-03-09"
format:
  revealjs:
    self-contained: true
    theme: dark  
    background-color: "black"  
    font-family: "Arial"  
    slide-number: true  
---

```{r include=FALSE, messages=FALSE, warning=FALSE}
library(tidyverse)
library(here)
# my_data <- read_csv(here("data", "my-data.csv"))
```

## Goals

- Identify Role-Based Age and Weight Trends: Use the scatterplot with regression lines to visualize how player age varies by weight across offense, defense, and special teams, revealing differences in career longevity and physical demands by role.

-Highlight Career Longevity vs. Injury Burden: Leverage the bubble plot to compare average career length and injury seasons by position, emphasizing how some roles (like kickers) are safer and more enduring than others (like tight ends).

-Expose Patterns Linking Physical Attributes to Risk: Together, the plots aim to uncover how physical characteristics and player roles correlate with career outcomes, helping explain why some athletes last longer or face greater injury risks.

## Data slide 1

- This analysis is based on seven .csv datasets related to NFL player careers and injuries. **Five** of the datasets provide season-by-season statistics for players at **key positions: ** quarterback, running back, wide receiver, defensive player, and kicker. The **sixth** dataset contains player background information, including **height, weight, and total years played** in the NFL. The **seventh** and final dataset includes **reported injury records** spanning several years.

- The data was collected by the **NFL** (National Football League) and its official analytics partner, **Next Gen Stats**, using comprehensive tracking and recording systems. These datasets are a massive sample of NFL players throughout the league’s history.

## Data slide 2

Show how many observations are in your dataset, the main variables that you are analyzing, and something about the scope of the data (levels of factors, years, range of variables)


## Weight of a Player vs Injuries Sustained

```{r echo=FALSE, out.height="50%", fig.align="center"}
mpg |> ggplot(aes(displ, cty)) + geom_point() + theme_bw() + theme(text=element_text(size=24))
Weight <- basic_stats$Weight..lbs

basic_stats_cleaned <- basic_stats %>%
  mutate(
    name_clean = str_to_lower(str_trim(str_replace(Name, "^(.*),\\s*(.*)$", "\\2 \\1"))),

    weight_char = as.character(Weight),  
    weight = as.numeric(str_extract(weight_char, "\\d+"))
  )

injury_counts <- injuries %>%
  mutate(name_clean = str_to_lower(str_trim(full_name))) %>%
  group_by(name_clean) %>%
  summarise(total_injuries = n())

weight_injury_data <- basic_stats_cleaned %>%
  left_join(injury_counts, by = "name_clean") %>%
  mutate(total_injuries = replace_na(total_injuries, 0)) %>%
  filter(!is.na(weight), weight >= 100)

ggplot(weight_injury_data, aes(x = weight, y = total_injuries)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  annotate("rect",
           xmin = 180, xmax = 270,
           ymin = 15, ymax = 60,
           fill = "red", alpha = 0.3, color = "darkred") +

  labs(
    title = "Player Weight vs Total Injuries",
    x = "Weight (lbs)",
    y = "Total Injuries"
  ) +
  theme_minimal()
```

## Average Career Length vs Average Seasons Injured by Position


position_summary <- basic_stats %>%
  mutate(
name_clean = str_to_lower(str_trim(str_replace(Name, "^(.*),\\s*(.*)$", "\\2 \\1"))),
    Experience_clean = ifelse(str_detect(Experience, "Rookie"), "0", Experience),
    career_length = suppressWarnings(parse_number(Experience_clean)),
    position_grouped = case_when(
      Position %in% c("CB", "S", "LB", "ILB", "OLB", "SS", "FS", "DT", "DE", "DB", "NT", "DL") ~ "DEF",
      Position %in% c("QB", "RB", "WR", "TE", "K") ~ Position,
      TRUE ~ NA_character_
    )
  ) %>%
  left_join(
    injuries %>%
      mutate(name_clean = str_to_lower(str_trim(full_name))) %>%
      distinct(name_clean, season) %>%
      group_by(name_clean) %>%
      summarise(injury_seasons = n()),
    by = "name_clean"
  ) %>%
  mutate(
    injury_seasons = replace_na(injury_seasons, 0),
    career_length = replace_na(career_length, 0)
  ) %>%
  filter(position_grouped %in% c("QB", "RB", "WR", "TE", "K", "DEF")) %>%
  group_by(position_grouped) %>%
  summarise(
    avg_career_length = mean(career_length),
    avg_injury_seasons = mean(injury_seasons),
    total_players = n()
  )

ggplot(position_summary, aes(x = avg_injury_seasons, y = avg_career_length, size = total_players, color = position_grouped)) +
  geom_point(alpha = 0.8) +
  scale_size_continuous(range = c(5, 15)) +
  labs(
    title = "Avg Injuries vs Career Length by Position",
    x = "Avg Injury Seasons per Player",
    y = "Avg Career Length (Years)",
    color = "Position",
    size = "Total Players"
  ) +
  theme_minimal()

## Influence of Weight on Injury Vulnerability

```{r}
all_data <- my_data$basic_stats

all_data <- my_data$basic_stats %>%
  mutate(
    name_clean = str_to_lower(str_trim(str_replace(Name, "^(.*),\\s*(.*)$", "\\2 \\1"))),
    position = Position,
    weight_lbs = as.numeric(str_extract(as.character(Weight), "\\d+")),
    age = as.numeric(str_extract(as.character(Age), "\\d+")),
    side_of_ball = case_when(
      position %in% c("QB", "RB", "FB", "WR", "TE", "OL", "C", "G", "T") ~ "Offense",
      position %in% c("DE", "DT", "LB", "OLB", "ILB", "MLB", "CB", "DB", "FS", "SS", "NT", "SAF", "EDGE") ~ "Defense",
      position %in% c("K", "P", "LS") ~ "Special Teams",
      TRUE ~ "Unknown"
    )
  ) %>%
  filter(
    !is.na(age), !is.na(weight_lbs),
    age < 45,
    weight_lbs >= 150, weight_lbs <= 400,
    side_of_ball != "Unknown"
  )

ggplot(all_data %>% filter(side_of_ball != "Unknown"), 
       aes(x = weight_lbs, y = Age, color = side_of_ball)) +
  geom_jitter(width = 5, height = 0.3, alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~ side_of_ball) +
  scale_color_manual(values = c("Offense" = "light green", "Defense" = "red", "Special Teams" = "blue")) +
  labs(
    title = "Age vs. Weight by Side of Ball",
    x = "Weight (lbs)",
    y = "Age",
    color = "Side of Ball"
  ) +
  theme_minimal()
```

## Conclusion

- Player roles strongly influence physical profiles and longevity—defensive players tend to be younger and heavier, while special teams players can maintain careers at older ages despite being lighter.

- Career length and injury risk vary widely by position—for example, kickers enjoy long, low-risk careers, while tight ends face the highest injury burden despite average longevity.

- Physical traits like weight play a key role in injury patterns—lighter players, especially in the 180–240 lb range, are more prone to injuries, particularly those in high-contact positions like wide receivers and running backs.


