---
title: "Give your project a title"
author: "Team and team member"
date: "2024-02-29"
output: html_document
---


<!-- Submission and Evaluation


Load this file in Rstudio and provide as much details as you can on your plan

When finished, submit your proposal in pdf and Rmd format to BS (Term project proposal)

Component	                    Score	Maximum
Questions	                    NA	    2
Data provided and read into R	NA	    1
Data description	            NA    	3
Visualization ideas          	NA    	3
References                   	NA    	1
Total	
10


Feedback


Think of the following sentences as a rubric that distinguishes between excellent, good, and insufficient proposals.

Your questions were easy to understand and of appropriate complexity. easy to understand but probably too ambitious for this project. easy to understand but too simple; try to think a bit more deeply about the data.

You provide your data in the data folder and showed how it could be read with R. provided your data in the data folder but did not show how it could be read with R. did not provide your data in the data folder.

You described the source, observations and variables in your data well enough so that I could make a connection between your data and your questions. You should provide more information about the source, observations, and variables in your data in your project report.

You described appropriate visualizations for your data and questions. Your plans for visualizations were not detailed enough for me to see the connection to your data or questions. You should think more creatively and deeply about ways to visualize your data as you work on your project.

You have a clear plan for the work of each team mate. You need to think more carefully about how you will divide the work of the project across your team.

You provided a reference so I know where your data came from. You did not provide reference for your data; what was its source? There was important technical information in your proposal which did not have a source. Please provide it in your report.

-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("ggfortify")
library(tidyverse)
library(here)
library(ggfortify)
library(broom)
library(ggrepel)
library(kableExtra)
library(ggplot2)
library(ggtext)
library(kableExtra)
library(dplyr)
library(knitr)
```

As you work on this proposal, write your responses under the corresponding heading. **Delete the instructions (including this paragraph) before submitting your proposal.**

## Questions to be investigated

Identify a 2-3 questions you'd like to investigate with your data visualizations. Provide enough background that the questions make sense to your reader.

## Data to be used

This is a key part of your proposal! 

Describe the data you plan to use including

* a general description of the dataset,
* the variables and type (categorical with a description of the levels, quantitative with units),
* the number of observations,
*  metadata that are relevant including who collected the data, when they were collected, how they were collected (a survey, random sample, exhaustive sample, experiment, etc.)

Include a copy of the data in one of three ways:
1. place your data file in your github repository and provide the link
2. provide a download link to your dataset (not necessarily on github)
3. or upload your dataset to BS with your proposal (preferred solution)

[In any case, I need to be able to access and use YOUR dataset.]

Read the data into R here. Present one or several summary tables or the output of `skimr::skim()` on your data to supplement the descriptive (in plain english not code) information you've provided.

No need to run any model here!

Look at the structure of the data. Explain the key variables that you plan to visualize/analyze (if your dataset has many, e.g. > 30 columns, you dont need to use all the columns).

```{r}
# my_data <- read_csv(...)
# Your data does not need to be in csv format.
```

## Ideas for visualization
Plan to use a PCA, to help reduce the complexity of the data and indentify paterns of the player performance over time. A regression model will help look at performance, injuries and age and how they influence player longevity. A cluster model will help group the players into different cluster based on their performance statistics and their longevity.

## Team planning
Both members working together in person on the same datatset, collaborting on each part of the project, no one individual is completing a section on their own.

## Challenges anticipated
No help needed.

## References

Describe the source of your data here, including a link to where the data can be found on the internet (or elsewhere!)

If you used any sources as inspiration give them as well. None are required.

