---
title: "NFL Player Longevity Trends and Insights"
author: "Lyle Huth and Callum Deveaux"
date: "2025-03-12"
output: html_document
---

<!-- Submission and Evaluation


Load this file in Rstudio and provide as much details as you can on your plan

When finished, submit your proposal in pdf and Rmd format to BS (Term project proposal)

Component	                    Score	Maximum
Questions	                    NA	    2
Data provided and read into R	NA	    1
Data description	            NA    	3
Visualization ideas          	NA    	3
References                   	NA    	1
Total	
10


Feedback


Think of the following sentences as a rubric that distinguishes between excellent, good, and insufficient proposals.

Your questions were easy to understand and of appropriate complexity. easy to understand but probably too ambitious for this project. easy to understand but too simple; try to think a bit more deeply about the data.

You provide your data in the data folder and showed how it could be read with R. provided your data in the data folder but did not show how it could be read with R. did not provide your data in the data folder.

You described the source, observations and variables in your data well enough so that I could make a connection between your data and your questions. You should provide more information about the source, observations, and variables in your data in your project report.

You described appropriate visualizations for your data and questions. Your plans for visualizations were not detailed enough for me to see the connection to your data or questions. You should think more creatively and deeply about ways to visualize your data as you work on your project.

You have a clear plan for the work of each team mate. You need to think more carefully about how you will divide the work of the project across your team.

You provided a reference so I know where your data came from. You did not provide reference for your data; what was its source? There was important technical information in your proposal which did not have a source. Please provide it in your report.

-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("ggfortify")
library(tidyverse)
library(here)
library(ggfortify)
library(broom)
library(ggrepel)
library(kableExtra)
library(ggplot2)
library(ggtext)
library(kableExtra)
library(dplyr)
library(knitr)
library(skimr)
```

As you work on this proposal, write your responses under the corresponding heading. **Delete the instructions (including this paragraph) before submitting your proposal.**

## Questions to be investigated

1. How does the position of an NFL player influence their career length and what factors may induce a shorter career?

2. How does the height and weight of a player at a given position affect their injury sensativity and comparatively, their willingness to take hits?

## Data to be used

* a general description of the dataset
We are analyzing data from seven different .csv datasets, five of which contain data of a player's career statistics by season, at each relevent position (being quarterback, running back, wide receiver, defensive player, and punter). The sixth dataset contains every players background information, including their height, weight, and years played in the NFL. The seventh and final dataset contains reported injury data over several years.

* the variables and type (categorical with a description of the levels, quantitative with units)
CATEGORICAL:
- Position (quarterback, running back, wide receiver, defensive player, punter)
- Injury designation (body part injured)
- Defensive play (sacks, interceptions, tackles)

QUANTITATIVE:
- passing yards + attempts
- rushing yards + attempts
- kicking yards + attempts
- Height (ft)
- Weight (lbs)
- career length (seasons - equivalent to years)
- length of injury (games - equivalent to weeks)

* the number of observations
Each dataset has between 8000 up to 80,000 observations. After cleaning the data, clearing out Null/ Nan values, we expect to have at least 1000 observations.

*  metadata that are relevant including who collected the data, when they were collected, how they were collected (a survey, random sample, exhaustive sample, experiment, etc.)
The data was collected by the nfl and official partner 'next gen stats' over the history of the national football league. The dataset takes the full population, but after cleaning if there is still an excess of data we may collect a smaller sample.

── Data Summary ────────────────────────
                           Values
Name                       data  
Number of rows             17172 
Number of columns          16    
_______________________          
Column type frequency:           
  character                12    
  numeric                  4     
________________________         
Group variables            None  

The categorical and quantitative variables used are above.

```{r}
file_list <- list.files(path = "2430_csv_files", pattern = "*.csv", full.names = TRUE)
my_data <- lapply(file_list, read.csv)

# Your data does not need to be in csv format.
```

## Ideas for visualization
We plan to use a PCA, to help reduce the complexity of the data and indentify patterns of a players performance over time. A regression model will help look at performance, injuries and age and how they influence player longevity. A cluster model will help group the players into different cluster based on their performance statistics and their longevity.

## Team planning
Our plan is that both members working together in person on the same datatset, collaborting on each part of the project, no one individual is completing a section on their own.
If we are intended to divide our work amongst eachother,
Lyle Huth:
- Plot construction of 2 of the 5 visualizations
- Data cleaning and prepping for analysis
- Sorting and choosing which variables to compare
- 

Callum Deveaux:
- Plot construction of 3 of the 5 visualizations
- Analysis of discovered trends
- Organization of our report and presentation

## Challenges anticipated
No help needed.

## References

Describe the source of your data here, including a link to where the data can be found on the internet (or elsewhere!)
We used kaggle.com for our datasets. We are aware of the risk of false or randomly generated data from kaggle, however the datasets used in this notebook were taken directly from the nfl.com official website.
https://www.kaggle.com/code/chastings25/nfl-career-health-and-length-analysis
If you used any sources as inspiration give them as well. None are required.

